---
title: "Intro to machine learning with R"
author: "Dereck de Mézquita"
date: "12/04/2020"
output:
  html_document: 
    number_sections: yes
    self_contained: no
    toc: yes
editor_options: 
  chunk_output_type: console
---

# Description

This .Rmd is associated to the `Intro to machine learning with R` course on derecksnotes.com. This is used to produce plots and other results for the course. It will be linked and published along with the course.

# Document set up

```{r}
knitr::knit2html()
```


```{r setup, include=FALSE}
par(mfrow = c(1, 1)); # Reseting margins on plots

knitr::opts_chunk$set(echo = TRUE, results = "hide", message = FALSE, warning = FALSE);
```

Here set the variables as described; these will determine how and what the programme executes.

```{r}
os <- "macOS"; # Select OS: macOS, linux, windows
wd <- "/Users/Work/Coding/projects-personal/web-sites/derecksnotes/courses/entries/documents/intro-ml-R" # dirname(rstudioapi::getActiveDocumentContext()$path); # If you want to set wd manually change this variable to a string of your path

setwd(wd)
```

# Linear regression
## History

In this example we try to draw a line as close to the every points that may appear on a plot. For classical linear regressior, aka "least squares method", you only need to measure the distance between the points on the Y axis.

This is simplist example possible.

Take the following code chunk and plot example.

### Example 1
```{r}
x <- 1:2;
y <- 1:2;

linearReg <- function(xs, ys) {
  fit <- lm(ys ~ xs) # Fitting linear model
  abline(fit, col = "red") # Adds straight line to plot; (y ~ x)
};

plot(x, y,
	 main = "Linear regression between two points",
	 xlim = c(0.75, 2.25),
	 ylim = c(0.75, 2.25));
linearReg(x, y);
```

### Example 2

In this example we do the regression test with the built in data set `iris`. This tracks flowers petal, sepal... etc. morphology per species.

This minimises the vertical distance between the points and the line. Here we take the lowess regression to get a best fit line to all points.

```{r}
lowessReg <- function(xs, ys) {
	fit <- lowess(xs, ys);
	lines(fit, col = "blue"); # lowess line (x,y)
};

plot(iris$Sepal.Length,
	 iris$Petal.Length,
	 main = "Sepal vs petal (length)");

lowessReg(iris$Sepal.Length, iris$Petal.Length);
```

Formulas for building linear regression models in R are built in the following way; `(y ~ x)`. In order to add more predictor variables to be taken into account use the `+` sign in the formula `(y ~ x + z)`.

Let's redo the above plot with a linear regression model as we first did but this time attributing more predictor variables.

```{r}
plot(iris$Sepal.Length,
	 iris$Petal.Length,
	 main = "Sepal vs petal (length) + sepal width");

abritrary_model <- lm(log(iris$Sepal.Length, base = 10) ~ iris$Petal.Length + iris$Sepal.Width, data = iris);

abline(abritrary_model, col = "purple");
```

Let's break down the formulas:

* `lm()`: a function for linear regression.

* `log(Sepal.Length, base = 10) ~ Petal.Length + Sepal.Width`: formula for modelling taken as input.

* `log(Sepal.Length, base = 10)`: the quantity we want to predict.

* `Petal.Length + Sepal.Width`: variables available to make predictions.

Once we have declared all those variables in the arbitrary model we can then use them for prediction:

```{r}
iris$predLog <- predict(abritrary_model, newdata = iris$Species)

iris$predLog
```


```{r}

```

